Supplementary material to: Drug efficacy prediction in cell lines

### 2019-01-10 docs/supplement.Rnw
simulation models now same framework, partly implemented in R

% R packages
<<echo=FALSE>>=
library(cambridge)
@

<<figures, cache=FALSE, echo=FALSE>>=
read_chunk("../code/R/figures.R")
@

% settings
\pgfplotsset{compat=1.13}
\graphicspath{{/Users/magnusmunch/Documents/OneDrive/PhD/project_cambridge/graphs/}}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Supplementary material to: Drug efficacy prediction in cell lines}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}

\begin{document}
	\maketitle

@@

	\section{Introduction}
	This document contains the Supplementary Material (SM) to the document `Drug efficacy prediction in cell lines'. In the following, this document is referred to as Main Document (MD).

@@

  \section{Model without tissue effects}
  \subsection{Introduction}
  In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$.

  \subsection{Model and estimation}
  The model is given by:
  \begin{align*}
  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
  \gamma_d^2 & \sim \mathcal{IG}(\theta_d, \lambda), \\
  \sigma_d^2 & \sim 1/\sigma_d^3.
  \end{align*}
  The corresponding variational Bayes posterior is given by:
  \begin{align*}
  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \frac{\lambda}{\theta_d^2}, \delta_d\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + p + 1}{2}, \zeta_d\).
  \end{align*}
  We update the variational parameters until convergence:
  \begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = \frac{2 \zeta_d^{(h)}}{n+p+1} (\X \tr \X + a_d^{(h)} \I)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + a_d^{(h)} \I)^{-1} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = \frac{n+p+1}{2 \zeta_d^{(h)}} [\trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}] + \lambda, \\ \zeta_d^{(h+1)} & = \frac{a_d^{(h+1)}}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace [ (\X \tr \X + \I_p) \bm{\Sigma}_d^{(h+1)}] + (\bm{\mu}_d^{(h+1)}) \tr (\X \tr \X + \I_p)\bm{\mu}_d^{(h+1)}\],
  \end{align*}
  with $a_d^{(h)}$ as in the full model in the MD. Equivalently, the empirical Bayes updates for $\bm{\alpha}$ and $\lambda$ are the same as in the full model in the MD.

  \subsection{Evidence lower bound}
  We monitor convergence through the evidence lower bound (ELBO):
  \begin{align*}
  \text{ELBO} & (Q^{(l)}) \propto \frac{1}{2} \sum_{d=1}^D \log |\bm{\Sigma}_d^{(l)}| - \frac{n + p + 1}{2} \sum_{d=1}^D \log \zeta_d^{(l)} + \sum_{d=1}^D \log K_{\frac{p+1}{2}}\(\sqrt{\lambda^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) \\
  & - \frac{n+p+1}{4} \sum_{d=1}^D (\zeta_d^{(l)})^{-1} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
  & + \frac{D(p+3)}{4} \log \lambda^{(l)} + \lambda^{(l)} \sum_{d=1}^D (\theta_d^{(l)})^{-1} - \frac{p+1}{2} \sum_{d=1}^D \log \theta_d^{(l)} - \frac{p+1}{4} \sum_{d=1}^D \log \delta_d^{(l)} \\
  & + \frac{1}{2} \sum_{d=1}^D \[ \delta_d^{(l)} - \lambda^{(l)} - \frac{n+p+1}{2\zeta_d^{(l)}} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - \frac{n+p+1}{2\zeta_d^{(l)}} \trace (\bm{\Sigma}_d^{(l)}) \] a_d^{(l)}. 
  \end{align*}

  - Different options for convergence. We may monitor parameters, ELBO, or a combination. We may also iterate a fixed number of times, especially for the VB (currently doing this, using 2 VB iterations and 20 EB iterations).

  \subsection{Efficient computation}
  The empirical Bayes updates require the following quantities: $\zeta_d$, $a_d$, $\delta_d$, $\trace (\bm{\Sigma}_d)$, $\trace (\X \tr \X \bm{\Sigma}_d)$, $\bm{\mu}_d \tr \bm{\mu}_d$, and $\bm{\mu}_d \tr \X \tr \X \bm{\mu}_d$. In addition, we need $\log|\bm{\Sigma}|$ and $\mathbf{y}_d \tr \X \bm{\mu}_d$ to monitor the ELBO. The first three quantities are obtained by scalar operations of $\mathcal{O}(n)$. The rest is easily calculated using the SVD $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$:
  \begin{align*}
  \trace (\bm{\Sigma}_d^{(h+1)}) & = \frac{2\zeta_d^{(h)}}{n + p + 1} \[\sum_{i=1}^n (v_i^2 + a_d^{(h)})^{-1} + \max (p - n, 0)\cdot (a_d^{(h)})^{-1}\], \\
  \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) & = \frac{2\zeta_d^{(h)}}{n + p + 1} \sum_{i=1}^n v_i^2(v_i^2 + a_d^{(h)})^{-1}, \\
  (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)})^2, \\
  (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^4 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)})^2, \\
  \log|\bm{\Sigma}^{(h+1)}| & \propto p \log \zeta_d^{(h)} - \sum_{i=1} \log (v_i^2 + a_d^{(h)}) - \max(p - n, 0) \cdot \log a_d^{(h)}, \\
  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} & = \sum_{i=1}^n v_i^2 [(\mathbf{U} \tr \mathbf{y}_d)_i]^2/(v_i^2 + a_d^{(h)}).
  \end{align*}
  where $v_i$ are the singular values of $\X$. The SVD and $\mathbf{U} \tr \mathbf{y}_d$ are $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ operations, respectively, but have to be calculated only once at the start of the algorithm. The rest of the calculations involve just $n$ scalar multiplications and/or additions.

  \subsection{Independent error variance model parameters}
  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}. However, in our case that is debatable. The $D$ regressions need not be on the same scale, so some form of calibration of the $\bm{\beta}_d$ posteriors to this scale is required. A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$. For a sense of completeness we still compare to the independent prior model in our simulations. The independent prior model is given by:
  \begin{align*}
  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
  \beta_{jd} | \gamma_d^2 & \sim \mathcal{N} (0, \gamma_d^2), \\
  \gamma_d^2 & \sim \mathcal{IG}((\mathbf{c}_d \tr \bm{\alpha})^{-1}, \lambda), \\
  \sigma_d^2 & \sim 1/\sigma_d^3.
  \end{align*}
  Its variational posterior is given by:
  \begin{align*}
  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+1}{2}, \lambda (\mathbf{c}_d \tr \bm{\alpha})^2, \delta_d\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{n + 1}{2}, \zeta_d\),
  \end{align*}
  with variational parameter updates:
  \begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = \(\frac{n+1}{2 \zeta_d^{(h)}}\X \tr \X + a_d^{(h)} \I_p\)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = \frac{n+1}{2 \zeta_d^{(h)}} \bm{\Sigma}_d^{(h+1)} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = \trace(\bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)} + \lambda, \\ \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace (\X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)}\],
  \end{align*}
  with $a_d^{(h)}$ and the empirical Bayes updates $\bm{\alpha}^{(l+1)}$ and $\lambda^{(l+1)}$ as in the full model in the MD. A singular value decomposition $\X = \mathbf{U} \mathbf{D} \mathbf{V} \tr$ leads to the following efficient computations:
  \begin{align*}
  \trace (\bm{\Sigma}^{(h+1)}_d) & = \frac{2\zeta^{(h)}_d}{n + 1} \sum_{i=1}^n \(v_i^2 + \frac{2\zeta^{(h)}_d a^{(h)}_d}{n+1}\)^{-1} + \max (p - n, 0)\cdot (a^{(h)}_d)^{-1}, \\
  \trace (\X \tr \X \bm{\Sigma}^{(h + 1)}_d) & = \frac{2\zeta^{(h)}_d}{n + 1} \sum_{i=1}^n v_i^2 \(v_i^2 + \frac{2\zeta^{(h)}_d a^{(h)}_d}{n+1}\)^{-1}, \\
  (\bm{\mu}_d^{(h + 1)}) \tr \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2 v_i^2 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\)^{-2}, \\
  (\bm{\mu}_d^{(h + 1)}) \tr \X \tr \X \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2v_i^4 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\)^{-2}, \\
  \log|\bm{\Sigma}^{(h + 1)}| & \propto p \log \zeta_d^{(h)} - \sum_{i=1} \log \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\) \\
  & \,\,\,\,\,\,\,\,\,\,\, - \max(p - n, 0) \cdot (\log a_d^{(h)} + \log \zeta_d^{(h)}), \\
  \mathbf{y}_d \tr \X \bm{\mu}_d^{(h + 1)} & = \sum_{i=1}^n [(\mathbf{U} \tr \mathbf{y}_d)_i]^2 v_i^2 \(v_i^2 + \frac{2 \zeta_d^{(h)} a_d^{(h)}}{n + 1}\),
  \end{align*}
  where $v_i$ are the singular values of $\X$. As before, this renders $\mathcal{O}(pn^2)$ and $\mathcal{O}(n^2)$ calculations once at the start of the algorithm and iterative updates of $\mathcal{O}(n)$.

  \subsection{Simulations}
  We simulate from a correctly specified model. We assume random $y_{id}$, $\x_i$, and $\bm{\beta}_d$, while we fix $\mathbf{C}$, $\gamma_d^2$, $\sigma_d^2$ and $\bm{\alpha}$. We estimate the $\bm{\mu}_d$, $\bm{\Sigma}_d$, $\delta_d$, $\bm{\alpha}$ and $\lambda$, while we fix the $\sigma_d^2$ to their true values.

  We set $n=100$, $p=200$, and $D=10$. We consider 5 evenly sized classes of drugs, such that $\mathbf{c}_{kd} = \mathbbm{1}(k = \text{class}_d)$ and $\begin{bmatrix} \mathbf{c}_1 & \cdots & \mathbf{c}_D \end{bmatrix} \tr$. From these the $\gamma_d^2 = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ are created. Next, we simulate $\bm{\beta}_{jd} \sim \mathcal{N} (0, \sigma^2_d \gamma_d^2)$. We fix the mean signal-to noise ratio $\overline{\text{SNR}}_d$, such that the predictor data variance is $s^2 = \V(\mathbf{x}_i) = \overline{\text{SNR}}_d / (\overline{\gamma^2_d} p)$ (NOT CORRECT). The last step consists of simulating the data by $\mathbf{x}_{ij} \sim \mathcal{N}(0,s^2)$ and $y_{id} = \mathcal{N} (\mathbf{x}_i \tr \bm{\beta}_d, \sigma_d^2)$. The remaining parameters are set as follows: $\bm{\alpha} = \begin{bmatrix} 1 & \dots & 5 \end{bmatrix} \tr$, $\sigma_d^2 = 1$ for all $d$, and $\overline{\text{SNR}}_d = 10$. To account for the random data and parameter generation, we repeat the simulation 100 times.

  We compare the results to a model where the $\gamma_d^2$ are drawn from a class-specific inverse Gamma distribution; a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}:
  \begin{align*}
  y_{id} | \bm{\beta}_d ,\sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
  \beta_{jd} | \gamma_d^2 ,\sigma_d^2 & \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2), \\
  \gamma_d^2 & \sim \Gamma^{-1} (a_{\text{class}(d)}, b_{\text{class}(d)}), \\
  \sigma_d^2 & \sim \Gamma^{-1} (0.001, 0.001),
  \end{align*}
  where $\text{class}(c)$ denotes the class of drug $d$. This allows for more flexible empirical Bayes estimation of the $a_d$ and $b_d$, and thus prior means of the different drug classes. As a consequence however, the risk of overfitting is increased. Additionally, it does not allow to include continuous covariates on the drugs. We present the results in Figures \ref{fig:boxplot_igaussian_res1_sigma2}-\ref{fig:boxplot_igaussian_res1_alpha}.

>>>>

% settings
\pgfplotsset{compat=1.13}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Supplementary material to: Drug efficacy prediction in cell lines}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: \href{mailto:m.munch@vumc.nl}{m.munch@vumc.nl}}, Mark A. van de Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}

\begin{document}

% load packages and set options
<<include=FALSE>>=
library(cambridge)
opts_knit$set(base.dir="../figs")
opts_chunk$set(fig.align='center', echo=FALSE, fig.path="../figs/")
@

% read figure code
<<figures, include=FALSE>>=
read_chunk("../code/figures.R")
@

	\maketitle

@@

	\section{Content overview}
	This document contains the Supplementary Material (SM) to the document `Drug efficacy prediction in cell lines'. In the following, this document is referred to as Main Document (MD).

@@

	\section{Simulations}
	\subsection{Models}
	In order to investigate the inverse Gaussian prior model and its estimation, we consider a simpler model. The simpler model ignores tissue effects and fixes $\forall g: \phi_g^2=1$:
	\begin{align*}
  y_{id} | \bm{\beta}_d, \sigma_d^2 & \sim \mathcal{N} (\x_i \tr \bm{\beta}_d, \sigma_d^2), \\
  \beta_{jd} | \gamma_d^2, \sigma_d^2 & \sim p(\beta_{jd}), \\
  \gamma_d^2 & \sim p(\gamma_d^2), \\
  \sigma_d^2 & \sim 1/\sigma_d^3.
  \end{align*}

  \cite{moran_variance_2018} argue that in regression models, $\bm{\beta}_d$ and $\sigma_d^2$ should be independent \textit{a priori}. However, in our case that is debatable. The $D$ regressions need not be on the same scale, so some form of calibration of the $\bm{\beta}_d$ posteriors to this scale is required. A simple way of doing this is to include the scales $\sigma_d^2$ in the priors for the $\bm{\beta}_d$. We call the two versions of the model conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ dependent):
  $$
  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \mathcal{N} (0, \sigma_d^2 \gamma_d^2)
  $$
  and non-conjugate (with $\bm{\beta}_d$ and $\sigma_d^2$ independent):
  $$
  \beta_{jd} | \gamma_d^2, \sigma_d^2 \sim \beta_{jd} | \gamma_d^2 \sim \mathcal{N} (0, \gamma_d^2),
  $$
  and compare them in our simulations.

	In addition to the inverse Gaussian model for the prior variances $\gamma_d^2$ (introduced in the MD):
	$$
  \gamma_d^2 \sim \mathcal{IG}(\theta_d, \lambda_d),
  $$
	we investigate a generalisation of the models in \cite{kpogbezan_empirical_2017} and \cite{leday_gene_2017}, where we draw the $\gamma_d^2$ from an inverse Gamma distribution; :
	$$
	\gamma_d^2 \sim \Gamma^{-1}(\alpha_d/2, \lambda_d/2).
	$$
	For the sake of comparability, we have rescaled the inverse Gamma parameters and reparametrised both distributions as generalized inverse Gaussians:
	\begin{align*}
  \gamma_d^2 & \sim \mathcal{GIG}(-1/2, \lambda_d \theta_d^{-2}, \lambda_d), \\
  \gamma_d^2 & \sim \mathcal{GIG}(-\alpha_d/2, 0, \lambda_d).
  \end{align*}
	The two models coincide if $\theta_d \to \infty$ and $\alpha_d = 1$.

	\subsection{Estimation}
	The variational distributions are as follows:
	\begin{align*}
  q(\bm{\beta}) & \overset{D}{=} \prod_{d=1}^D \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \\
  q(\bm{\gamma}^2) & \overset{D}{=} \prod_{d=1}^D \mathcal{GIG}\(-\frac{p+\alpha_d}{2}, \lambda_d \theta_d^{-2}, \delta_d\), \\
  q(\bm{\sigma}^2) & \overset{D}{=} \prod_{d=1}^D \Gamma^{-1} \(\frac{df + 1}{2}, \zeta_d\),
  \end{align*}
	where $\alpha_d=1$ and $\theta_d \to \infty$ in the inverse Gaussian and inverse Gamma models, respectively. Furthermore, $df=n + p$ in the conjugate model and $df=n$ in the non-conjugate model. The variational parameters contain cyclic dependencies and are, in the conjugate setting, iteratively updated by:
	\begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} \I)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} \I)^{-1} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = a_d^{(h)} \bmu_d \tr \bmu_d + \lambda_d, \\ 
  \zeta_d^{(h+1)} & = \frac{1}{2} \bigg[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \\
  & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + b_d^{(h+1)} \trace ( \bm{\Sigma}_d^{(h+1)}) + b_d^{(h+1)} (\bm{\mu}_d^{(h+1)}) \tr \bm{\mu}_d^{(h+1)}\bigg].
  \end{align*}
	while in the non-conjugate setting, we use the following updating equations:
	\begin{align*}
  \bm{\Sigma}_d^{(h+1)} & = (a_d^{(h)})^{-1} (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1}, \\
  \bm{\mu}_d^{(h+1)} & = (\X \tr \X + b_d^{(h)} (a_d^{(h)})^{-1} \I)^{-1} \X \tr \y_d, \\
  \delta_d^{(h+1)} & = \bmu_d \tr \bmu_d + \lambda_d, \\ 
  \zeta_d^{(h+1)} & = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -2 \mathbf{y}_d \tr \X \bm{\mu}_d^{(h+1)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(h+1)}) + (\bm{\mu}_d^{(h+1)}) \tr \X \tr \X \bm{\mu}_d^{(h+1)} \].
  \end{align*}
  Here, we set
  \begin{align}
  a_d^{(h)} & =\E_{Q^{(h)}}(\sigma_d^{-2})=(df+1)/(2 \zeta_d^{(h)}), \nonumber \\
  b_d^{(h)} & = \E_{Q^{(h)}}(\gamma_d^{-2}) = \sqrt{\frac{\lambda_d}{\theta_d^2 \delta_d^{(h)}}} \frac{K_{\frac{p + \alpha_d -2}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)}{K_{\frac{p+\alpha_d}{2}} \( \sqrt{\lambda_d \delta_d^{(h)}}/\theta_d \)} + \frac{p+\alpha_d^{(h)}}{\delta_d^{(h)}}. \label{eq:auxvar1}
  \end{align}
  Note that in the inverse Gamma model, the computation of (\ref{eq:auxvar1}) simplifies to $b_d^{(h)}=(p+\alpha_d)/\delta_d^{(h)}$.

  \subsection{Empirical Bayes}
  To avoid subjectivity in our choice of prior, we estimate the hyper-parameters with empirical Bayes. The empirical Bayes procedure differs for the two prior models. In the inverse Gaussian case, we parametrise $\theta_d = (\mathbf{c}_d \tr \bm{\alpha})^{-1}$ to allow for the inclusion of drug covariates in the empirical Bayes estimation. The estimating equations are:
  \begin{align*}
  \bm{\alpha} & = \[ \mathbf{C} \tr \diag (\lambda_d e_d^{(l)}) \mathbf{C} \]^{-1} \diag (\lambda_d) \mathbf{1}_{D \times 1}, \\
  \lambda_d & = \[ b_d^{(l)} + e_d^{(l)} (\mathbf{c}_d \tr \bm{\alpha})^2 - 2\mathbf{c}_d \tr \bm{\alpha} \]^{-1},
  \end{align*}
  where $e_d^{(l)} = (b_d^{(l)} + \delta_d^{(l)}/p) \cdot \delta_d^{(l)} (\theta_d^{(h)})^2/\lambda_d^{(l)}$. Solving the equations is done by iteratively reweighted least squares of responses $(e_d^{(l)})^{-1}$ on predictors $\mathbf{c}_d$ with weights $\lambda_d e_d^{(l)}$. If we use one common $\lambda=\lambda_d$ then the solution is in closed form:
  \begin{align*}
	\bm{\alpha}^{(l+1)} & = \[ \mathbf{C} \tr \diag(e_d^{(l)}) \mathbf{C} \]^{-1} \mathbf{C} \tr \mathbf{1}_{D \times 1}, \\
	\lambda^{(l+1)} & = \[ \sum_{d=1}^D b_d^{(l)} - (\bm{\alpha}^{(l+1)}) \tr \mathbf{C} \tr \mathbf{1} \]^{-1} D.
	\end{align*}
	This corresponds to the model in the MD.

  In the inverse Gamma model we use the following empirical Bayes procedure: Suppose we have a partitioning of our drugs into drug classes and we let prior variance components $\gamma_d^2$ and $\gamma^2_{d'}$ be drawn from the same distribution if drug $d$ and drug $d'$ are from the same class. Or simply put: $\gamma^2_{d}, \gamma^2_{d'} \sim \Gamma^{-1}(\alpha_c/2, \lambda_c/2)$ if $d, d' \in \text{class}_c$. Empirical bayes estimation of the $\alpha_c$ and $\lambda_c$ by maximisation of the marginal likelihood amounts to iteratively solving the estimating equations:
  \begin{subequations}
    \begin{align}
      \log \(\frac{\lambda_c}{2}\) - \psi\(\frac{\alpha_c}{2}\) - \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} & = 0, \label{eq:estequation1}\\
      \frac{\alpha_c}{\lambda_c} - \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} & = 0, \label{eq:estequation2}
    \end{align}
  \end{subequations}
  where $e^{(l)}_d = \log \lambda_d^{(l)} - \psi ( \alpha_d^{(l)}/2) - \log 2$. We solve (\ref{eq:estequation2}) for $\lambda_c$, plug it into (\ref{eq:estequation1}) and bound the result using \cite{alzer_inequalities_1997} to obtain an interval $( a_c^*, 2a_c^* )$ containing the solution to (\ref{eq:estequation1}), where:
  $$
  a_c^* = \left\{ \frac{\sum_{d \in \text{class}_c} e^{(l)}_d}{|\text{class}_c|} + \log \[ \frac{\sum_{d \in \text{class}_c} b_d^{(l)}}{|\text{class}_c|} \]  \right\}^{-1}.
  $$
  With this interval it is straightforward to find $\alpha^{(l+1)}_c$ by any root-finding algorithm and plugging the solution into (\ref{eq:estequation2}) to find $\lambda^{(l+1)}_c$.

	The inverse Gamma model with independent drug classes allows for more flexible empirical Bayes estimation of the prior means for the different drug classes. As a consequence however, the risk of overfitting is increased. Additionally, it does not allow to include continuous covariates on the drugs.

	\subsection{Evidence lower bound}
	We monitor convergence through the evidence lower bound (ELBO). To compute the evidence lower bound after updating the EB parameters from $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}$ to $\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$, we compute $\text{ELBO}^{(l + 1)} = \sum_{d=1}^D \text{ELBO}_d^{(l + 1)}$, with, in the conjugate case:
	\begin{align*}
	\text{ELBO}_d^{(l + 1)} & = \frac{1}{2} \log |\bm{\Sigma}_d^{(l)}| - \frac{a_d^{(l)}}{2} \[ \y_d \tr \y_d - 2\y_d \tr \X \bm{\mu}_d^{(l)} + (\bm{\mu}_d^{(l)}) \tr \X \tr \X \bm{\mu}_d^{(l)} + \trace ( \X \tr \X \bm{\Sigma}_d^{(l)})\] \\
	& - \frac{df + 1}{2} \log \zeta_d^{(l)} + \frac{p + \alpha_d^{(l)}}{4} \log \lambda_d^{(l)} - \frac{p + \alpha_d^{(l)}}{2} \log \theta_d^{(l)} - \frac{p+\alpha_d^{(l)}}{4} \log \delta_d^{(l)} \\
	& + \log K_{\frac{p+\alpha_d^{(l)}}{2}}\(\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}\) + \frac{b_d^{(l)}}{2} \[ \delta_d^{(l)} - a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)} - a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)}) \] \\
	& + \frac{\lambda_d^{(l)} e_d^{(l)}}{2(\theta_d^{(l)})^2} + \frac{(\alpha_d^{(l)} - \alpha_d^{(l + 1)})h_d^{(l)}}{2} - \log K_{\frac{\alpha_d^{(l + 1)}}{2}}\(\lambda_d^{(l + 1)}/\theta_d^{(l + 1)}\) + \frac{\alpha_d^{(l + 1)}}{2} \log \theta_d^{(l)} \\
	& - \frac{\lambda_d^{(l+1)} e_d^{(l)}}{2 (\theta_d^{(l + 1)})^2} - \frac{\lambda_d^{(l + 1)} b_d^{(l)}}{2},
	\end{align*}
	where $h_d^{(l)} = \log \delta_d^{(l)} - \psi [(p + \alpha_d^{(l)})/2] - \log 2$ in the inverse Gamma model and
	$$
	h_d^{(l)} = \frac{1}{2}\log \delta_d^{(l)} - \frac{1}{2}\log \lambda_d^{(l)} + \log \theta_d^{(l)} + \frac{\partial}{\partial \nu} \log K_{\nu} (\sqrt{\lambda_d^{(l)} \delta_d^{(l)}}/\theta_d^{(l)}) \bigg|_{\nu = \frac{p + \alpha_d^{(l)}}{2}}
	$$
	in the inverse Gaussian case. Note that in the inverse Gaussian case, the computationally expensive $h_d^{(l)}$ terms never have to be calculated, since we have $\alpha_d^{(l)} - \alpha_d^{(l + 1)}=0$. The non-conjugate ELBO simply exchanges the $a_d^{(l)} (\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $a_d^{(l)} \trace (\bm{\Sigma}_d^{(l)})$ terms with $(\bm{\mu}_d^{(l)}) \tr \bm{\mu}_d^{(l)}$ and $\trace (\bm{\Sigma}_d^{(l)})$, respectively. To compute the ELBO after a VB update we simply let $\alpha_d^{(l)}, \theta_d^{(l)}, \lambda_d^{(l)}=\alpha_d^{(l + 1)}, \theta_d^{(l + 1)}, \lambda_d^{(l + 1)}$.

  In practice, there are different options to assess convergence. We may monitor the parameters themselves, the ELBO, or a combination of the two. We may also iterate for an \textit{a priori} fixed number of times. We may also choose to use a different convergence criterium for the EB and VB iterations. Currently we fix the number of EB iterations to 20, while we do 2 VB iterations per EB iteration.
